\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2014-01-30 fluB.tex
Adding text revisions suggested by @rambaut
Text revisions, moving MDS plots to supplementary info and using them
in supplementary info.

\author[1]{Gytis Dudas}
\author[2]{Trevor Bedford}
\author[1]{Samantha Lycett}
\author[1,3]{Andrew Rambaut}

\affil[1]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
\affil[2]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
\affil[3]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}

@@

\section*{Introduction}
Seasonal influenza causes an estimated 250,000 to 500,000 deaths annually and is comprised of three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality.
Though lacking the degree of antigenic diversity that influenza A viruses possess, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.

Like all members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments (a process known as reassortment). 
Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of seasonal human and non-human influenza A strains. 
The only known persistent host of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000,bodewes2013}), thus limiting the virus's potential to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir. 
Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2013}.

Currently circulating influenza B viruses are comprised of two distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after strains B/Victoria/02/87 and B/Yamagata/16/88, that are thought to have genetically diverged in HA around 1983 \cite{rota1990}. 
These two lineages possess antigenically distinct HA surface glycoproteins allowing them to co-circulate in the human population.
All other influenza B segments also show a split into two lineages around that time.

Previously, through the combined use of time-stamped and maximum likelihood phylogenetic trees evolutionary rate, selective pressures and reassortment history of major clades of influenza B have been analyzed \cite{chen2008}, showing extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
This analysis was limited because it used limited and different numbers of sequences to infer trees of each segment and relied on maximum likelihood phylogenies to infer the history of reassortments.
Here, we introduce another approach that reveals a evolutionarily intriguing pattern of reassortment in influenza B.
In this approach, membership to either the Victoria or Yamagata lineage in the tree of one segment is used to label the individual viruses in the tree of the other segments.
Reassortment events which result in the replacement of one segment lineage by another show up as changes in the discrete trait along a branch (Figure \ref{methodFig}).
Here, we use discrete traits to reconstruct major reassortment events in the recent history of influenza B viruses and to partition diversity across segments of influenza B virus strains to quantify reassortment dynamics over time.

@@

We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012. 
Each strain was assigned to a lineage (either Vic or Yam) in each segment by making maximum likelihood trees and looking whether it fell within the subtree containing B/Victoria/02/87 or B/Yamagata/16/88 sequences with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.

@@

\subsection*{Building trees}
Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
Here, we modeled the substitution process using the HKY model of nucleotide substitution, with separate HKY models for each of the 3 codon partitions, and additionally included robust counting \cite{obrien2009} to report realized substitutions.
We modeled the demographic process using a GMRF Bayesian skyride model \cite{minin2008}.
We used `relaxed' tips for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
We ran 3 independent MCMC chains, each with 200 million states, sampled every 20000 steps and discarded the first 10\% of the MCMC states as burn-in.
Samples were combined across runs to give 27000 total samples for each segment and convergence was assessed by Tracer.

@@

\subsection*{Measures of diversity}
We inferred the diversity of each segment over time by finding the most recent common ancestor of all branches existing at yearly intervals, which places an upper boundary on the maximum amount of diversity existing at each time period.
In addition, we calculated mean pairwise TMRCA between branches inferred to be in virions possessing Vic and Yam lineage PB1, PB2 and HA.
This gave us an averaged measure of how much a particular segment reassorts with respect to PB1, PB2 and HA segments.
If Vic and Yam lineages of PB1, PB2 and HA segments were to be considered as being separate populations this measure would be equivalent to `between population' diversity.

We also calculated the total amount of evolutionary time each segment has spent in virions with entirely Vic, entirely Yam or mixed lineage PB1, PB2 and HA segments by taking the sum of branch lengths in each tree under different combinations of the PB1, PB2 and HA labels.
This gives a measure of how successful, over long periods of time, each particular PB1-PB2-HA constellation has been.

\subsection*{Tree to tree similarities}
In order to get a measure of tree to tree similarity we first extracted 900 trees from the posterior distributions of trees for each segment.
These were combined across 3 independent MCMC chains to give a total of 2700 trees.
Pairwise comparison between each pair of segments was performed by comparing trees at each MCMC state.

@@

If we are comparing TMRCAs of the same two tips in tree A and tree B we express it as $\frac{[A:A']+[B:B']}{2\times [A:B]}$, where [A:B] is $\Delta$TMRCA of a pair of tips between tree A and tree B and [A:A'] is $\Delta$TMRCA between the same pair of tips between tree A and a replicate of tree A (A'), to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.

We also attempted to calculate subtree prune and regraft (SPR) distances between trees using rSPR \cite{whidden2009,whidden2010,whidden2013}.
However, because the time taken to calculate the SPR distances depends on the SPR distances themselves, we find that exact SPR distance calculations over posterior sets of trees are impractical for all but the most similar trees.
We thus calculated approximate SPR distances for all trees, using the normalization procedure described above.

\subsection*{Estimating linkage disequilibrium across the influenza B genome}
We attempted to estimate linkage disequilibrium (LD) between amino acid sites across the influenza B proteome.
We chose the $\chi^{2}_{df}$ statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
LD was estimated only at loci where each `allele' was represented by at least two isolates and gaps were ignored and not considered as polymorphisms.
Ignoring gaps in estimating LD was a practical decision based on some sequences missing nucleotides at the beginning and end of sequences.
However, there is at least one polymorphic indel within the HA protein, which is associated with the Vic-Yam split and not with sequence quality.
In addition, some influenza B genome constellations have not reassorted for prolonged periods of time and have acquired amino acid substitutions on several segments, resulting in linkage disequilibrium.
These single site pairs result in high LD and come from poorly sampled lineages, thus we choose to only focus on alleles that have been sampled at least twice.

We also calculated mean LD for all pairs of segments to quantify whole-segment associations.

\subsection*{Confirming findings with a larger dataset through lineage assignment}
The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates sampled 1984-2013 were downloaded from GISAID.
Neighbour-joining trees of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/02/87 or B/Yamagata/16/88 sequences.
Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.

@@

The split of Vic and Yam lineages can be seen in all segments \cite{chen2008}.
Following the split of the two lineages, each segment could be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
Some viruses with mixed-lineage genomes have become common ancestors to many later strains (i.e. the `trunk' of the phylogenetic tree) and the segment lineages involved so successful, as to become fixed within the influenza B population circulating today.

We find that Vic lineage PA, NP, NA and MP segments have gone extinct (\textit{i.e.} no Vic lineage sequences of those segments have been sampled in recent years see Figure \ref{lineageRatiosOverTime}), replaced entirely by the respective segments of Yam lineage in viruses bearing a Vic lineage HA.
Similarly, Vic lineage NS has become fixed in the influenza B virus population as well.
By looking at the most recent common ancestor of all lineages existing at yearly timeslices within the tree of each segment (see Figure \ref{tmrcaOT}), it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, whereby all sequences sampled more recently trace their ancestry to increasingly more recent common ancestors.

\subsection*{PB1, PB2 and HA segments continue to circulate as two lineages}
We find that PB1, PB2 and HA segments of influenza B have not undergone lineage turnover, unlike PA, NP, NA, MP and NS segments.
This is shown in Figure \ref{tmrcaOT}, where PB1, PB2 and HA lineages are shown to be derived from a consistently old TMRCA, indicating continued accumulation of diversity since the split of Vic and Yam lineages.
We suggest that the maintenance of both Vic and Yam lineages in PB1, PB2 and HA segments might be caused by the action of some form of balancing selection.
In Figure \ref{lineageRatiosOverTime} the ratios at which Vic and Yam lineages of PB1, PB2 and HA segments have been isolated appear to be close to 0.5 over long periods of time.
On a year-to-year basis the ratios can fluctuate dramatically and suggesting frequency-dependent domination of one or the other lineage within a given year.

@@

We focus only on PB1, PB2 and HA lineage labels (traits), since all other segments have fixed either the Vic or the Yam lineage.
Figure \ref{betweenDiversity} shows that all trees have reassorted with respect to Victoria and Yamagata lineages of PB1, PB2 and HA.
However, we also see that PB1, PB2 and HA traits show reciprocal preservation of diversity after 1997.
This suggests that no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments (i.e. PB1, PB2 and HA traits after 1997 only `meet' at the root of PB1, PB2 and HA trees).
In a time period close to the initial split of Vic and Yam lineages (1986--1996) we see reduced diversity between Vic and Yam labels of PB1 and PB2 segments on one hand and HA on the other.
These reductions in diversity represent small clades with PB1 and PB2 segments from one lineage and the HA segment from another, which go extinct by 1997.

@@

We find that PB1, PB2 and HA segments, in addition to not reassorting across the Vic-Yam lineage boundary since 1997, form co-reassorting segment complexes derived entirely of Vic or Yam lineage segments.
Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations.

We have identified 3 classes of mixed-lineage PB1-PB2-HA reassortants from our data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), YVV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and YVY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
An additional class of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated June 9 -- November 12 in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
In all cases PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
In particular reassortment events combining PB1 and PB2 segments of different lineages, \textit{e.g.} B/Nanchang/630/1994-like and B/New York/24/1993-like isolates (each having two sequences), exhibit poor sampling and short circulation times (up to two years in the case of B/Nanchang/630/1994-like viruses).
In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bankok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past (up to 4 years for B/Bankok/163/1990-like, although B/Waikato/6/2005-like only circulated for 5 months).

@@

We reconstructed reassortment events detectable by discrete traits.
Figure \ref{railroadPlot} focuses only on reassortments that have occured after 1990.
We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS) circulating 1992 -- 2011: B/Alaska/12/1996-like (YYYYYYYV), B/Nanchang/2/1997-like (VVYVYVYV), B/Iowa/03/2002-like (VVYVYYYV), B/California/NHRC0001/2006-like (VVYVYYYV) and B/Brisbane/33/2008-like (VVYVYYYV).
Figure \ref{railroadPlot} shows that reassorting segments appear to evolve with a considerable degree of autonomy.
For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to B/Iowa/03/2002-like isolates.
A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has been derived from Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.

@@

\subsection*{PB1, PB2 and HA trees experience reassortment `distance' effects}
Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of trees after normalization (see Methods).
If there are biases in the way segments reassort, so that some segments tend to co-reassort more often, we expect to see small-scale similarities between phylogenetic trees of those segments.
In our case we expect SPR distances (which are proportional to the numbers of reassortment that have taken place between trees) to reflect the raw reassortment rate.
In Figure \ref{SPR distances} the 95\% HPD intervals of normalized approximate SPR distances between pairs of segments overlap, suggesting there is no evidence of differences in the numbers of reassortments (or reassortment rate given as number of SPR moves per total time in the tree, data not shown) between all segments.
We note that our SPR distance analysis might simply lack power.
SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
In addition, we use approximate SPR distances, rather than exact SPR distances.
However, we do estimate exact SPR distances for a limited number of segment pairs (PB1-PB2 and PB1-HA) and find that after normalization exact and approximate SPR distances are not significantly different.

Although we do not find evidence of differences in numbers of raw reassortments between all pairs of trees, we find support for a reassortment `distance' effect.
Figure \ref{deltaTMRCA} shows normalized mean $\Delta$TMRCA values for all pairs of trees.
Most segment pairs show very low values for this statistic, indicating differences in TMRCAs.
PB1, PB2 and HA trees, on the other hand, exhibit TMRCAs between themselves which are, on average, much more similar to the replicates of those trees.
This suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves with lineages with which they have not recently shared the same virion.
In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees.
Previously, associations between MP and NA segments have been suspected in influenza A viruses (Samantha Lycett, personal communication).

@@

\subsection*{Linkage disequilibrium analysis confirms phylogenetic findings}
Linkage disequilibrium analyses (Figure \ref{segmentLD}) confirm our findings from phylogenetic analyses that PB1, PB2 and HA segments are linked.
We observe much higher mean $\chi^{2}_{df}$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
In some cases this value exceed the estimated LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA).
Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting reciprocally high LD.
These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
In addition, there are sites on PB1, PB2 and HA proteins which also show high (albeit smaller) LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.

\tbc{The 8x8 MDS analysis suggests that PB1-PB2-HA are the only segments with substantial coevolution.}

\begin{figure}[h]
	\centering	
	\includegraphics[width=0.75\textwidth]	{figures/InfB_segmentLD.png}
	\caption{\textbf{LD comparison between influenza B proteins.}
Comparison of linkage disequilibrium between amino acid sites on influenza B proteins.
Numbers on the heatmap are mean $\chi^{2}_{df}$ between pairs of proteins.}
	\label{segmentLD}
\end{figure}

>>>>

\author[1]{Gytis Dudas}
\author[2]{Trevor Bedford}
\author[1,3]{Samantha Lycett}
\author[1,4]{Andrew Rambaut}

\affil[1]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
\affil[2]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
\affil[3]{Institute of Biodiversity Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK}
\affil[4]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}

@@

\section*{Introduction}
Seasonal influenza causes an estimated 250,000 to 500,000 deaths annually and is comprised of three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality.
Although the known antigenic diversity of influenza B viruses is less than that of influenza A viruses, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.

Like all members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments (a process known as reassortment). 
Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of seasonal human and non-human influenza A strains. 
Influenza B viruses are thought to primarily infect humans (and occasionaly seals \cite{osterhaus2000,bodewes2013}), and are thought to be unable to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir. 
Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2013}.

Currently circulating influenza B viruses comprise two distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after strains B/Victoria/02/87 and B/Yamagata/16/88, that are thought to have genetically diverged in HA around 1983 \cite{rota1990}. 
These two lineages possess antigenically distinct HA surface glycoproteins allowing them to co-circulate in the human population.
All other influenza B segments also show a split into two lineages around that time.

Previously, through phylogenetic analyses the evolutionary rate, selective pressures and reassortment history of major clades of influenza B have been analyzed \cite{chen2008}, showing extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
Here, we expand upon the methods used previously and introduce another approach that reveals a evolutionarily intriguing pattern of reassortment in influenza B.
In this approach, membership to either the Victoria or Yamagata lineage in the tree of one segment is used to label the individual viruses in the tree of the other segments.
By modelling these labels on a phylogenetic tree, reassortment events which result in the replacement of one segment lineage by another show up as label changes along a branch (Figure \ref{methodFig}).
We use this method to reconstruct major reassortment events in the recent history of influenza B viruses and to partition diversity across segments of influenza B virus strains to quantify reassortment dynamics over time.

@@

We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in supplementary information). 
The segments of each strain were assigned to a lineage (either Vic or Yam) by making maximum likelihood trees of each segment and identifying whether the isolate was more closely related to B/Victoria/02/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.

@@

\subsection*{Building trees}
Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
Here, we modeled the substitution process using the HKY model of nucleotide substitution, with separate HKY models for each of the 3 codon partitions, and additionally used a recently described method for reporting estimates of realized synonymous and non-synonymous substitutions \cite{obrien2009}.
We used a flexible demographic model (Bayesian skyride \cite{minin2008}).
We accounted for incomplete sampling dates for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
We ran 3 independent MCMC chains, each with 200 million states, sampled every 20000 steps and discarded the first 10\% of the MCMC states as burn-in.
After assessing convergence of all 3 MCMC chains by Tracer, we combined all analyses to give a total of 27000 samples from the posterior distribution of trees.

@@

\subsection*{Measures of diversity}
We inferred the diversity of each segment over time by plotting the most recent common ancestor of all branches existing at yearly intervals, which places an upper bound on the maximum amount of diversity existing at each time period.
In addition, we calculated mean pairwise TMRCA between branches labelled as Vic and Yam for PB1, PB2 and HA.
This gave us a measure of how much a particular segment reassorts with respect to PB1, PB2 and HA segments.
If Vic and Yam lineages of PB1, PB2 and HA segments were to be considered as being separate populations this measure would be equivalent to `between population' diversity.

We also calculated the total amount of evolutionary time spent by each segment with entirely Vic, entirely Yam or mixed lineage PB1, PB2 and HA segments.
We do this by summing the branch lengths in each tree under 3 different lineage combinations of the PB1, PB2 and HA segments: PB1-PB2-HA derived entirely from Yamagata lineage, PB1-PB2-HA entirely derived from Victoria lineage and PB1-PB2-HA derived from a mixture of the two lineages.
This gives a measure of how successful, over long periods of time, each particular PB1-PB2-HA constellation has been.

\subsection*{Tree to tree similarities}
We subsampled our combined posterior distribution of trees to give a total of 2700 trees.
Pairwise comparison between each pair of segments was performed by comparing trees at each MCMC state.

@@

If we are comparing TMRCAs of the same two tips in tree A and tree B we express it as $\frac{[A\_{i}:A\_{i}']+[B\_{i}:B\_{i}']}{2\times [A\_{i}:B\_{i}]}$, where [A\_{i}:B\_{i}] is $\Delta$TMRCA of a pair of tips between \textit{i}th tree derived from alignment A and \textit{i}th tree derived from alignment B and [A\_{i}:A\_{i}'] is $\Delta$TMRCA between the same pair of tips between \textit{i}th tree derived from alignment A and \textit{i}th tree derived from an independent run of alignment A, to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.

Subtree prune and regraft (SPR) distances between phylogenetic trees are an approximate measure of the numbers of reassortment or recombination events.
Exact SPR distances are difficult to compute, as they depend on the SPR distance itself and are impractical to compute for posterior distributions of trees except for the most similar trees.
We calculated approximate SPR distances \cite{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
Approximate SPR distances were normalized using the procedure described above.

\subsection*{Estimating linkage disequilibrium across the influenza B genome}
We attempted to estimate linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
We use the $\chi^{2}_{df}$ statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
LD was estimated only at loci where each `allele' was represented by at least two isolates.
We ignored gaps in the alignment and did not consider them as polymorphisms.
However, there is at least one polymorphic indel within the HA protein, which is associated with the Vic-Yam split and not with incomplete sequences.

We also calculated mean LD for all pairs of segments to quantify the overall association between them.

\subsection*{Confirming findings with a larger dataset through lineage assignment}
The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates sampled 1984-2013 were compiled from GISAID.
Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/02/87 or B/Yamagata/16/88 sequences.
Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.

@@

The split of Vic and Yam lineages can be seen in all segments \cite{chen2008}.
Following the split of the two lineages, each segment could be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
Some viruses with mixed-lineage genomes have become fixed in the influenza B virus population \textit{i.e.} became the `trunk' of the phylogenetic tree of a segment.

We find that Vic lineage PA, NP, NA and MP segments have gone extinct (\textit{i.e.} no Vic lineage sequences of those segments have been sampled in recent years see Figure \ref{lineageRatiosOverTime}), replaced entirely by the respective segments of Yam lineage in viruses bearing a Vic lineage HA.
Conversely, Vic lineage NS has become fixed in the influenza B virus population as well.
By looking at the most recent common ancestor of all lineages existing at yearly timeslices within the tree of each segment (see Figure \ref{tmrcaOT}), it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, whereby all sequences sampled more recently trace their ancestry to increasingly more recent common ancestors.

\subsection*{PB1, PB2 and HA segments continue to circulate as two lineages}
We find that PB1, PB2 and HA segments of influenza B have not undergone lineage turnover, unlike PA, NP, NA, MP and NS segments.
This is shown in Figure \ref{tmrcaOT}, where PB1, PB2 and HA lineages are shown to be derived from a consistently old TMRCA, indicating continued accumulation of diversity since the split of Vic and Yam lineages.
We suggest that the maintenance of both Vic and Yam lineages in PB1, PB2 and HA segments might be caused by the action of some form of balancing selection.
In Figure \ref{lineageRatiosOverTime} the ratios at which Vic and Yam lineages of PB1, PB2 and HA segments have been isolated appear to be close to 0.5 over long periods of time.
On a year-to-year basis the ratios can fluctuate dramatically suggesting frequency-dependent domination of one or the other lineage within a given year.

@@

We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.
Increasing mean between label TMRCAs for all segments in Figure \ref{betweenDiversity} indicate that every segment has reassorted with respect to Victoria and Yamagata lineages of PB1, PB2 and HA segments.
However, we also see that PB1, PB2 and HA labels show reciprocal preservation of diversity after 1997.
This suggests that no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments (\textit{i.e.} PB1, PB2 and HA traits after 1997 only `meet' at the root of PB1, PB2 and HA trees).
In a time period close to the initial split of Vic and Yam lineages (1986--1996) we see reduced diversity between Vic and Yam labels of PB1 and PB2 segments on one hand and HA on the other.
These reductions in diversity represent small clades with PB1 and PB2 segments from one lineage and the HA segment from another, which go extinct by 1997.

@@

We find that PB1, PB2 and HA segments, in addition to not reassorting across the Vic-Yam lineage boundary since 1997, form co-reassorting segment complexes derived entirely of Vic or Yam lineage segments.
Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations (Figure \ref{stateTime}).

We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), YVV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and YVY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated June 9 -- November 12 in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
In all cases, PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
In particular reassortment events combining PB1 and PB2 segments of different lineages, \textit{e.g.} B/Nanchang/630/1994-like and B/New York/24/1993-like isolates (each having two sequences), exhibit poor sampling and short circulation times (up to two years in the case of B/Nanchang/630/1994-like viruses).
In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bankok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past (up to 4 years for B/Bankok/163/1990-like, although B/Waikato/6/2005-like only circulated for 5 months).

@@

We reconstructed reassortment events that were detected by using lineage labels.
Figure \ref{railroadPlot} focuses only on reassortments that have occured after 1990.
We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS) circulating 1992 -- 2011: B/Alaska/12/1996-like (YYYYYYYV), B/Nanchang/2/1997-like (VVYVYVYV), B/Iowa/03/2002-like (VVYVYYYV), B/California/NHRC0001/2006-like (VVYVYYYV) and B/Brisbane/33/2008-like (VVYVYYYV).
Figure \ref{railroadPlot} shows that reassorting segments appear to evolve with a considerable degree of autonomy.
For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to B/Iowa/03/2002-like isolates.
A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has been derived from Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.

@@

\subsection*{PB1, PB2 and HA trees experience reassortment `distance' effects}
Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of trees after normalization (see Methods).
If there are biases in the way segments reassort, so that some segments tend to co-reassort more often, we expect to see small-scale similarities between phylogenetic trees of those segments.
In our case we expect SPR distances (which are proportional to the numbers of reassortment that have taken place between trees) to reflect the raw reassortment rate.
In Figure \ref{SPRdistances} the 95\% HPD intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments (or reassortment rate given as number of SPR moves per total time in the tree, data not shown) between all segments.
We note that our SPR distance analysis might simply lack power.
SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
In addition, we use approximate SPR distances, rather than exact SPR distances.
However, we do estimate exact SPR distances for a limited number of segment pairs (PB1-PB2 and PB1-HA) and find that after normalization exact and approximate SPR distances are not significantly different.

Although we do not find evidence of differences in numbers of raw reassortments between all pairs of trees, we find support for a reassortment `distance' effect.
Figure \ref{deltaTMRCA} shows normalized mean $\Delta$TMRCA values for all pairs of trees.
Most segment pairs show very low values for this statistic, indicating differences in TMRCAs.
PB1, PB2 and HA trees, on the other hand, exhibit TMRCAs between themselves which are, on average, much more similar to the replicates of those trees.
This suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves unless both reassorting segments have similar TMRCAs, \textit{i.e.} there is isolation by distance.
In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees.
Previously, associations between MP and NA segments have been suspected in influenza A viruses (Samantha Lycett, personal communication).

@@

\subsection*{Linkage disequilibrium analysis confirms phylogenetic findings}
Although not an independent validation of association between PB1, PB2 and HA segments, linkage disequilibrium analyses show that Victoria and Yamagata lineages of these segments have accumulated lineage-specific amino acid substitutions.
In Figure \ref{segmentLD} their signature is greater mean $\chi^{2}_{df}$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
In some cases this value exceeds the estimated LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA).
Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting reciprocally high LD (see Supplementary information).
These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
In addition, there are sites on PB1, PB2 and HA proteins which also show high (albeit smaller) LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.

\begin{figure}[h]
	\centering	
	\includegraphics[width=0.75\textwidth]	{figures/InfB_aaMeanLD.png}
	\caption{\textbf{LD comparison between influenza B proteins.}
Comparison of linkage disequilibrium between amino acid sites on influenza B proteins.
Numbers on the heatmap are mean $\chi^{2}_{df}$ between pairs of proteins.}
	\label{segmentLD}
\end{figure}
